Towards Practical First-Order Model Counting

### 2024-04-19 doc/paper.tex
Section 2.2

\subsection{Algebra (TODO)}\label{sec:algebra}

\begin{itemize}
  \item We write $\expr{}$ for an arbitrary algebraic expression.
  \item Some terms have different meanings in the context of algebra as compared
        to logic. Here, a \emph{constant} is a non-negative integer. Similarly,
        a \emph{variable} is either a parameter of a function or a variable
        introduced by a summation, e.g., $i$ in $\sum_{i=1}^{n} \expr$.
\end{itemize}

\begin{definition}
  A (function) \emph{signature} is $f(x_{1}, \dots, x_{n})$ (written
  $f(\mathbf{x})$ for short), where $f$ is an $n$-ary function, and each $x_{i}$
  is a variable.
\end{definition}

\begin{definition}
  An \emph{equation} is $f(\mathbf{x}) = \expr{}$, where $f(\mathbf{x})$ is a
  signature. We refer to $f(\mathbf{x})$ and $\expr{}$ as the \emph{left-hand
    side} (LHS) and the \emph{right-hand side} (RHS) of the equation,
  respectively.
\end{definition}

\begin{definition}\label{def:basecase}
  Let $f$ be a function with signature $f(\mathbf{x})$. Then a \emph{base case}
  of $f$ is a function call $f(\mathbf{y})$ such that either $y_{i} = x_{i}$ or
  $y_{i} \in \mathbb{N}_{0}$ for all $i$, and the latter case applies at least
  once.
\end{definition}

@@

\begin{theorem}[Halting]\label{thm:halting}
  Let us assume the following about the input set of equations $\mathcal{E}$:
  \begin{enumerate}
    \item Each function $f$ has exactly one equation with $f$ on the LHS.\@ We
          call this equation the \emph{definition} of $f$.
    \item There exists a topological ordering of all functions ${(f_{i})}_{i}$
          such that the definition of $f_{i}$ does not contain function calls to
          $f_{j}$ with $j > i$.\footnote{This condition excludes the possibility
          of mutual recursion and similar cyclic scenarios and is akin to
          stratified logic programs~\cite{DBLP:books/sp/Lloyd87}.}
    \item For every equation $(f(\mathbf{x}) = \expr) \in \mathcal{E}$, every
          recursive function call $f(\mathbf{y}) \in \expr$ satisfies the
          following:
          \begin{itemize}
            \item each $y_{i}$ is either $x_{i} - c_{i}$ or $c_{i}$ for some
                  constant $c_{i} \in \mathbb{N}_{0}$;
            \item there exists $i$ such that $y_{i} = x_{i} - c_{i}$ for some
                  $c_{i} > 0$.
          \end{itemize}
  \end{enumerate}
  Then the set of base cases $\mathcal{B}$ returned by
  \FindBaseCases{$\mathcal{E}$} is \emph{sufficient} for $\mathcal{E}$ in the
  following sense. The evaluation of any function call with non-negative
  arguments halts as long as the evaluation of base cases is prioritised over
  the corresponding recursive definitions.\footnote{Recall that, as previously
    discussed, the order in which base cases are considered is immaterial.}
\end{theorem}

@@

>>>>

\subsection{Algebra}\label{sec:algebra}

We write $\expr{}$ for an arbitrary algebraic expression. Note that some terms
have different meanings in the context of algebra as compared to logic. Here, a
\emph{constant} is a non-negative integer. Similarly, a \emph{variable} is
either a parameter of a function or a variable introduced by a summation, e.g.,
$i$ in $\sum_{i=1}^{n} \expr$. A (function) \emph{signature} is
$f(x_{1}, \dots, x_{n})$ (written $f(\mathbf{x})$ for short), where $f$ is an
$n$-ary function, and each $x_{i}$ is a variable. An \emph{equation} is
$f(\mathbf{x}) = \expr{}$, where $f(\mathbf{x})$ is a signature.

\begin{definition}
  Let $f$ be a function with signature $f(\mathbf{x})$. Then a \emph{base case}
  of $f$ is a function call $f(\mathbf{y})$ such that either $y_{i} = x_{i}$ or
  $y_{i} \in \mathbb{N}_{0}$ for all $i$, and the latter case applies at least
  once.
\end{definition}

@@

\begin{theorem}[Halting]\label{thm:halting}
  Let us assume the following about the input set of equations $\mathcal{E}$:
  \begin{enumerate}
    \item Each function $f$ has exactly one equation with $f$ on the left-hand
          side.\@ We call this equation the \emph{definition} of $f$.
    \item There exists a topological ordering of all functions ${(f_{i})}_{i}$
          such that the definition of $f_{i}$ does not contain function calls to
          $f_{j}$ with $j > i$.\footnote{This condition excludes the possibility
          of mutual recursion and similar cyclic scenarios and is akin to
          stratified logic programs~\cite{DBLP:books/sp/Lloyd87}.}
    \item For every equation $(f(\mathbf{x}) = \expr) \in \mathcal{E}$, every
          recursive function call $f(\mathbf{y}) \in \expr$ satisfies the
          following:
          \begin{itemize}
            \item each $y_{i}$ is either $x_{i} - c_{i}$ or $c_{i}$ for some
                  constant $c_{i} \in \mathbb{N}_{0}$;
            \item there exists $i$ such that $y_{i} = x_{i} - c_{i}$ for some
                  $c_{i} > 0$.
          \end{itemize}
  \end{enumerate}
  Then the set of base cases $\mathcal{B}$ returned by
  \FindBaseCases{$\mathcal{E}$} is \emph{sufficient} for $\mathcal{E}$ in the
  following sense. The evaluation of any function call with non-negative
  arguments halts as long as the evaluation of base cases is prioritised over
  the corresponding recursive definitions.\footnote{Recall that, as previously
    discussed, the order in which base cases are considered is immaterial.}
\end{theorem}

@@

\todo[inline]{Remove mentions of LHS and RHS.}
